% \VignetteEngine{knitr::rmarkdown}
% \VignetteIndexEntry{"mtable" results in HTML format}

```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(comment=NA,
               fig.align="center",
               results="markup")
```

---
title: A demonstration of showing 'mtable' results in HTML format
---

# A demonstration of showing `mtable` results in HTML format

First, we load the package and estimate some regression models.

```{r, message=FALSE}
library(memisc)
lm0 <- lm(sr ~ pop15 + pop75,              data = LifeCycleSavings)
lm1 <- lm(sr ~                 dpi + ddpi, data = LifeCycleSavings)
lm2 <- lm(sr ~ pop15 + pop75 + dpi + ddpi, data = LifeCycleSavings)
```

Next, we put them together into a table with `mtable()`.

```{r}
mtable123 <- mtable("Model 1"=lm0,"Model 2"=lm1,"Model 3"=lm2,
                    summary.stats=c("sigma","R-squared","F","p","N"))
```
Third, we improve the appearance of the results by relabeling the `mtable` object.
```{r, comment=NA}
mtable123 <- relabel(mtable123,
                      "(Intercept)" = "Constant",
                      pop15 = "Percentage of population under 15",
                      pop75 = "Percentage of population over 75",
                      dpi = "Real per-capita disposable income",
                      ddpi = "Growth rate of real per-capita disp. income"
)
```
Finally we view the results in HTML format.
```{r, results='asis'}
show_html(mtable123)
```
The results can be written into an HTML file using
`write_html(mtable123,file="mtable123.html")` or the like. MS Word or LibreOffice can import such a file. 

[^1]: Unfortunately, LibreOffice currently seems to do a less than perfect job in importing HTML tables. The columns are a bit too wide and need to be adjusted by hand as do the skips after paragraphs. The example you find at the link mentioned above is thus manually adjusted.

